### 2021-11-08 Acta Informatica2022/1_intro.tex
Acta Informatica format

Satisfiability of word equations is a special instance of Hilbert’s 10th problem. 
In the past, the original motivation of studying word equations was for finding an undecidability proof of Hilbert’s 10th problem. 
However, the motivation is no longer valid since Makanin finds a decision procedure. Recently, driven by the need for program analysis, people started to revisit the problem and its extensions to describe the complete string library APIs in conventional programming languages. Many highly efficient solvers for string constraints, to name a few, CVC4~\cite{cvc4Tool}, Z3~\cite{z3}, Z3-Str3~\cite{zheng2013z3}, S3~\cite{trinh2014s3}, Norn~\cite{abdulla2015norn}, Ostrich~\cite{chen2017decidable}, Sloth~\cite{sloth}, ABC~\cite{aydin2018parameterized}, Stranger~\cite{yu2010stranger}, and Trau~\cite{abdulla2018trau}, are developed in the last decade.

The satisfiability of \emph{string-integer conversion constraints}, e.g., $wx=yx \wedge |x| > \parseInt(y)$, has been proven undecidable in~\cite{DayGHMN18}. 
However, this kind of constraints is pervasive in scripting language programs. 
For example, it is common that programs read string inputs from text files and converts a part of the string input to integers.
Even more crucially, in many programming languages, the string-integer conversion is a part of the definition of their core semantics~\cite{ecmascript2019ecmascript}. 
JavaScript, which powers most interactive content on the Web and increasingly server-side code with Node.js, is one of such languages.

@@

In this paper, we give a complete solution to this problem.
We propose a decision procedure for the class of word equation + flat languages + length constraints + string-integer conversion.
The basic idea of our approach can be sketched as follows: we first reduce the satisfiability problem to the corresponding satisfiability problem of the theory of Presburger Arithmetic with exponential functions (denoted by $\paexp$), \wuhao{more precisely, the existential fragment of $\paexp$}; then, according to the decidability of $\paexp$, we obtain the decidability of the original satisfiability problem.

@@

Aiming at introducing Point's procedure to computer science researchers, we reformulate (and slightly improve) Point's procedure in a way, which, hopefully, is more accessible for computer science researchers (Section~\ref{sec-dec}). This can be seen as another contribution of this paper.
\wuhao{We also analyze the deterministic upper bound of complexity for the first time: given a $\paexp$ formula with only existential quantifiers, eliminating all quantifiers has a 3-EXPSPACE complexity}. Furthermore, we propose various optimizations (Section~\ref{sec-opt}) and achieve the first prototypical implementation of Point's procedure (Section~\ref{sec-exp}).

>>>>

Satisfiability of word equations is a special instance of Hilbert’s 10th problem. 
In the past, the original motivation of studying word equations was for finding an undecidability proof of Hilbert’s 10th problem. 
However, the motivation is no longer valid since Makanin finds a decision procedure. Recently, driven by the need for program analysis, people started to revisit the problem and its extensions to describe the complete string library APIs in conventional programming languages. Many highly efficient solvers for string constraints, to name a few, CVC4~\cite{cvc4Tool}, Z3~\cite{z3}, Z3-Str3~\cite{zheng2013z3}, S3~\cite{trinh2014s3}, Norn~\cite{abdulla2015norn}, Ostrich~\cite{chen2017decidable}, Sloth~\cite{sloth}, ABC~\cite{aydin2018parameterized}, Stranger~\cite{yu2010stranger}, and Trau~\cite{abdulla2018trau}, are developed in the last decade.
The satisfiability of \emph{string-integer conversion constraints}, e.g., $wx=yx \wedge |x| > \parseInt(y)$, has been proven undecidable in~\cite{DayGHMN18}. 
However, this kind of constraints is pervasive in scripting language programs. 
For example, it is common that programs read string inputs from text files and converts a part of the string input to integers.
Even more crucially, in many programming languages, the string-integer conversion is a part of the definition of their core semantics~\cite{ecmascript2019ecmascript}. 
JavaScript, which powers most interactive content on the Web and increasingly server-side code with Node.js, is one of such languages.

@@

In this paper, we give a complete solution to this problem.
We propose a decision procedure for the class of word equation + flat languages + length constraints + string-integer conversion.
The basic idea of our approach can be sketched as follows: we first reduce the satisfiability problem to the corresponding satisfiability problem of the theory of Presburger Arithmetic with exponential functions (denoted by $\paexp$), more precisely, the existential fragment of $\paexp$; then, according to the decidability of $\paexp$, we obtain the decidability of the original satisfiability problem.

@@

Aiming at introducing Point's procedure to computer science researchers, we reformulate (and slightly improve) Point's procedure in a way, which, hopefully, is more accessible for computer science researchers (Section~\ref{sec-dec}). This can be seen as another contribution of this paper.
We also analyze the deterministic upper bound of complexity for the first time: given a $\paexp$ formula with only existential quantifiers, eliminating all quantifiers has a 3-EXPSPACE complexity. Furthermore, we propose various optimizations (Section~\ref{sec-opt}) and achieve the first prototypical implementation of Point's procedure (Section~\ref{sec-exp}).

### 2021-11-08 Acta Informatica2022/2_preliminary.tex
Acta Informatica format

An \emph{alphabet} $\Sigma$ is a finite set. Elements of $\Sigma$ are called \emph{letters}.
%Let $\Sigma$ denote a finite alphabet,
A \emph{string} $w$ over $\Sigma$ is a (possibly empty) finite sequence $a_1\ldots a_n$ with $a_i \in \Sigma$ for every $i \in [n]$. Let $\varepsilon$ denote the empty string, namely, the empty sequence. For a string $w = a_1 \ldots a_n \in \Sigma^*$, let $|w|$ denote the \emph{length} of $w$, i.e. $n$. In particular, $|\varepsilon| = 0$.
%
For $w_1 = a_1 \ldots a_m, w_2 = b_1 \ldots b_n\in \Sigma^*$, 
let $w_1\cdot w_2$ denote the \emph{concatenation} of $w_1$ and $w_2$, that is, $a_1 \ldots a_m b_1 \ldots b_n$.
%
Let $\Sigma^*$ denote the set of all strings over $\Sigma$ and $\Sigma^+$ denote the set of nonempty strings over $\Sigma$. 
For convenience, we also use $\Sigma_{\epsilon}$ to denote $\Sigma \cup \{\epsilon\}$.
%
A language $L$ over $\Sigma$ is a subset of $\Sigma^*$.

\paragraph{Presburger Arithmetic} \label{PA}
A Presburger Arithmetic ({\pa}) formula 
% is a first order theory over the signature 
%$\Sigma_\mathbb{N}\Def  \{0,1,+, <, |\}$. 
 is defined by the rules 
 $$
 \begin{array}{ l c l}
 \iterm &\Def & c \mid \ivarx \mid \iterm+\iterm \mid \iterm - \iterm, \\
 \phi & \Def & \iterm \ \op\ \iterm \mid c | \iterm \mid \phi \wedge \phi \mid \phi \vee \phi \mid \neg \phi \mid \exists \ivarx.\ \phi \mid \forall \ivarx.\ \phi,

 \end{array}
 $$
where $\op \in \{=, < , >, \le, \ge\}$ and $\ivarx, c$ are integer variables and constants, respectively. A \emph{quantifier-free} {\pa} (\qfpa) formula is a {\pa} formula containing no quantifiers. A {\pa} formula is in \emph{negation normal form} (NNF) if all occurrences of the negation symbol $\neg$ are before the atomic formulas. A {\pa} formula is called \emph{existential} if it is in NNF and contains no occurrences of universal quantifiers.
%An \emph{existential} LIA (ELIA) formula is an LIA formula where there are no universal quantifiers and all existential quantifiers are in the scope of an even number of the negation symbols $\neg$. 
%
The set of free variables of $\phi$, denoted by $\free(\phi)$, is defined in a standard manner. 
We usually write $\phi(\ivarx_1,\cdots, \ivarx_k)$ to denote an PA formula $\phi$ such that $\free(\phi) \subseteq \{\ivarx_1,\cdots, \ivarx_k\}$. Given an PA formula $\phi$, and an integer interpretation of $\free(\phi)$, i.e. a function $I: \free(\phi) \rightarrow \Int$, we denote by $I \models \phi$ that $I$ satisfies $\phi$ (which is defined in the standard manner, with $+$, $-$, and $|$ interpreted as the integer addition, subtraction, and divisibility relation respectively), and call $I$ a \emph{model} of $\phi$. We use $\llbracket \phi \rrbracket$ to denote the set of models of $\phi$.

>>>>

An \emph{alphabet} $\Sigma$ is a finite set. Elements of $\Sigma$ are called \emph{letters}.
%Let $\Sigma$ denote a finite alphabet,
A \emph{string} $w$ over $\Sigma$ is a (possibly empty) finite sequence $a_1\ldots a_n$ with $a_i \in \Sigma$ for every $i \in [n]$. Let $\varepsilon$ denote the empty string, namely, the empty sequence. For a string $w = a_1 \ldots a_n \in \Sigma^*$, let $|w|$ denote the \emph{length} of $w$, i.e. $n$. In particular, $\vert \varepsilon\vert  = 0$.
%
For $w_1 = a_1 \ldots a_m, w_2 = b_1 \ldots b_n\in \Sigma^*$, 
let $w_1\cdot w_2$ denote the \emph{concatenation} of $w_1$ and $w_2$, that is, $a_1 \ldots a_m b_1 \ldots b_n$.
%
Let $\Sigma^*$ denote the set of all strings over $\Sigma$ and $\Sigma^+$ denote the set of nonempty strings over $\Sigma$. 
For convenience, we also use $\Sigma_{\epsilon}$ to denote $\Sigma \cup \{\epsilon\}$.
%
A language $L$ over $\Sigma$ is a subset of $\Sigma^*$.

\paragraph{Presburger Arithmetic} \label{PA}
A Presburger Arithmetic ({\pa}) formula 
% is a first order theory over the signature 
%$\Sigma_\mathbb{N}\Def  \{0,1,+, <, |\}$. 
 is defined by the rules 
 $$
 \begin{array}{ l c l}
 \iterm &\Def & c \mid \ivarx \mid \iterm+\iterm \mid \iterm - \iterm, \\
 \phi & \Def & \iterm \ \op\ \iterm \mid c \vert \iterm \mid \phi \wedge \phi \mid \phi \vee \phi \mid \neg \phi \mid \exists \ivarx.\ \phi \mid \forall \ivarx.\ \phi,
 \end{array}
 $$
where $\op \in \{=, < , >, \le, \ge\}$ and $\ivarx, c$ are integer variables and constants, respectively. A \emph{quantifier-free} {\pa} (\qfpa) formula is a {\pa} formula containing no quantifiers. A {\pa} formula is in \emph{negation normal form} (NNF) if all occurrences of the negation symbol $\neg$ are before the atomic formulas. A {\pa} formula is called \emph{existential} if it is in NNF and contains no occurrences of universal quantifiers.
%An \emph{existential} LIA (ELIA) formula is an LIA formula where there are no universal quantifiers and all existential quantifiers are in the scope of an even number of the negation symbols $\neg$. 
%
The set of free variables of $\phi$, denoted by $\free(\phi)$, is defined in a standard manner. 
We usually write $\phi(\ivarx_1,\cdots, \ivarx_k)$ to denote an PA formula $\phi$ such that $\free(\phi) \subseteq \{\ivarx_1,\cdots, \ivarx_k\}$. Given an PA formula $\phi$, and an integer interpretation of $\free(\phi)$, i.e. a function $I: \free(\phi) \rightarrow \Int$, we denote by $I \models \phi$ that $I$ satisfies $\phi$ (which is defined in the standard manner, with $+$, $-$, and $\vert$ interpreted as the integer addition, subtraction, and divisibility relation respectively), and call $I$ a \emph{model} of $\phi$. We use $\llbracket \phi \rrbracket$ to denote the set of models of $\phi$.
